---
title: "housing"
format: pdf
execute:
  echo: false
  warning: false
---

## Housing

### The Irish Housing Crisis - Quick Analysis with CSO Data

```{r}
library(csodata)
library(dplyr)
library(ggplot2)
library(cowplot)
source("theme.R")
```

```{r, cache=TRUE}
new_dwelling_completions_by_type_and_quarter = csodata::cso_get_data(
  "NDQ01", use_factors = FALSE, pivot_format = "tidy"
)
```

```{r}
cso_rename_func <- function(col_name) {
  new_name <- stringr::str_replace_all(col_name, "\\.", "_")
  new_name <- tolower(new_name)
  new_name <- stringr::str_replace_all(new_name, " ", "_")
  new_name
}

new_dwelling_completions_by_type_and_quarter = rename_with(
  new_dwelling_completions_by_type_and_quarter, .fn = cso_rename_func
)
```

```{r}
new_dwelling_completions_by_type_and_quarter <- 
  new_dwelling_completions_by_type_and_quarter |> 
  mutate(quarter_date = lubridate::yq(quarter),
         year = stringr::str_remove(quarter, "Q.*"))

new_dwelling_completions_by_type_and_quarter
```

#### Definitions

-   [**Single**: If a single domestic dwelling or farm premises is to be connected to the ESB Network, the dwelling is defined as 'single'](https://www.cso.ie/en/releasesandpublications/ep/p-ndc/newdwellingcompletionsq22024/backgroundnotes/)

-   [**Scheme**: If a new multi-unit development with two or more houses is to be connected to the ESB Network, each dwelling is defined as a 'scheme'](https://www.cso.ie/en/releasesandpublications/ep/p-ndc/newdwellingcompletionsq22024/backgroundnotes/)

-   [**Apartment**: If a new multi-unit development with two or more apartments is to be connected to the ESB Network, each dwelling is defined as 'apartment'](https://www.cso.ie/en/releasesandpublications/ep/p-ndc/newdwellingcompletionsq22024/backgroundnotes/)

```{r}
NEW_DWELLING_REQUIRED_ESTIMATE <- 50000

new_dwelling_completions_by_type_and_year <-
  new_dwelling_completions_by_type_and_quarter |> 
  group_by(year, type_of_house) |> 
  summarise(new_dwelling_completions = sum(new_dwelling_completions)) 

house_type_stacked_p = new_dwelling_completions_by_type_and_year |> 
  filter(type_of_house != "All house types") |> 
  ggplot(
    aes(
      year, 
      new_dwelling_completions, 
      group = type_of_house, 
      fill = type_of_house
    )
  ) + 
  geom_col() +
  geom_hline(
    yintercept = NEW_DWELLING_REQUIRED_ESTIMATE, 
    colour = ukraine_palette$ukraine_blue,
    linewidth = 0.7,
    linetype = 2
  ) +
  annotate(
    "text",
    x = "2021",
    y = NEW_DWELLING_REQUIRED_ESTIMATE - 1500,
    label = "Estimate of number of dwellings required per year",
    colour = ukraine_palette$ukraine_blue_dark
  ) +
  ukraine_plot_theme() +
  scale_fill_manual(values = c(
    ukraine_palette$ukraine_blue,
    ukraine_palette$ukraine_yellow_darkened,
    ukraine_palette$ukraine_blue_dark
    )
  ) +
  labs(title = "Ireland - Dwellings Built per Year",
       x = "Year",
       y = "# Dwellings Completed")

house_type_filled_p <- 
  new_dwelling_completions_by_type_and_year |> 
  filter(type_of_house != "All house types") |> 
  ggplot(
    aes(
      year, 
      new_dwelling_completions, 
      group = type_of_house, 
      fill = type_of_house
    )
  ) + 
  geom_col(position= "fill") +
  ukraine_plot_theme() +
  scale_fill_manual(values = c(
    ukraine_palette$ukraine_blue,
    ukraine_palette$ukraine_yellow_darkened,
    ukraine_palette$ukraine_blue_dark
    )
  )

house_type_stacked_p 
house_type_filled_p
```
